---
title: "Introduction to tidycensus"
output: html_notebook
---

```{r loading, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidycensus)

# insert your census api key here # This works but it seems to take a little longer for it to work, so keep trying with the following commands if it gives you an error, it is a brick wall but very breakable
ckey <- readLines("~/Documents/GitHub/ADJHomework/census_key.txt")
census_api_key(ckey)
```

Here's the documentation for [tidycensus](https://walker-data.com/tidycensus/articles/basic-usage.html#geography-in-tidycensus-1)

Start with a very basic request: population by county
```{r}
county_pop <- get_acs(geography = "county", #differentiate by county
              variables = "B01001_001", #population variable
              state = "MO", #specify state
              year = 2019) #specify most recent year
```

Now by place (and you can rename the variable): 
```{r}
place_pop <- get_acs(geography = "place", #place is more specific than county as there are many places in a county
              variables = c(population="B01001_001"), #calling the pop value to be population so we don't have a stroke while we read
              state = "MO", 
              year = 2019)
```

Search for variables:
```{r}
acs2019_var <- load_variables(2019, "acs5", cache=TRUE)

acs2019_var %>% filter(grepl("income", label))
```

Also, you can use data.census.gov to locate the variables you're interested in.
(Search for "plumbing")
```{r}
acs2019_var %>% filter(grepl("B25049", name))
```


Try for some more variables (plumbing facilities)
```{r}
plumbing <- get_acs(geography = "place", #want to keep it by cities
              variables = c("B25049_001","B25049_002","B25049_003","B25049_004","B25049_005","B25049_006","B25049_007"), #all of the plumbing variables
              state = "MO", 
              year = 2019)
```
### CENSUS 2020 ###

Only redistricting data is currently available from the 2020 Census.
If you read the documentation on the tidycensus website, it says that the dataset name for the redistricting data is "pl".

Load the variables file for the redistricting data: 
```{r}

```

Get population by "place" in Missouri:
```{r}

```

More granularly, get population by tract in Boone County:
```{r}

```

Compare to 2010 Census
First get the 2010 population numbers by tract in Boone County:
```{r}

```

Then join the two datasets together:
```{r}

```
